Master thesis problem statement: Keyword search over heterogeneous sources

### 2012-10-11 doc/thesis/latex/report_sept.tex
updated problem definition, solutions and literature review. review to be finished

Using a highly structured language like SQL is problematic because users need to know not only the language but also where to find the information and also lots of technical details like schema. A data integration system based on simple structured queries is already in place. Still {\color{red}various improvements including support} for less restricted keyword queries and improvements to system's usability and performance still have to be researched.

@@

\section{Case study: the CMS Experiment at CERN}
Users' information need may vary greatly depending on their role, however most of the time they are interested in locating a \textit{full set} of entities matching some selection criteria, e.g.:
%\footnote{in contrary to exploratory approach where  a subset of best matches is enough}
   \begin{itemize}
  		\item find all \textit{files} from \textit{dataset(s)} matching wild-card query each containing some of the 'interesting' \textit{runs} from a list provided (Release validation teams)
         \item find (all) \textit{datasets} related some specific physics phenomena\footnote{In case of dataset this data is present in filename or run} together with conditions describing how this data was recorded by detector or simulated which are present in separate autonomous system than the datasets (Physicists)
         % TODO:is this a good example? conditions is a separate query now...
         \item find all \textit{datasets} matching some pattern stored at a given \textit{site} (filtering on entities stored on separate services)
   \end{itemize}                		
% In such dynamic environment it is preferred to choose local-as-view approach for information integration instead of creating a global schemata\cite{Koch00CERN}.

@@

The Data Aggregation System (DAS)\cite{Kuznetsov2010, Kuznetsov2011} was created which allows integrated access to a number of proprietary data-sources by processing user's queries on demand - it queries the data-sources, merges the results, and caches them for subsequent use. DAS uses \textit{Boolean retrieval model} as users are often interested in retrieving ALL the items matching their query.

Currently the queries specify what entity the user is interested in (dataset, file, etc) and provide selection criteria (attribute=value, name BETWEEN [v1, v2]) operators. The combined query results could be later 'piped' for further filtering and aggregation (min, avg, etc), e.g.:

{\footnotesize 
\begin{verbatim}
dataset=*RelVal* | grep dataset.nevents >1000 | avg(dataset.size), median(dataset.size)
\end{verbatim}
}

@@

\section{Problem statement}

@@

Therefore, it is important to balance between returning locally cached query results quickly   and between issuing queries to slow data-services to get up to date results.
It is also not less important to provide users with easy to use interface having a fairly flat learning curve, while at the same time not loosing the functionality of fairly complex queries. % expressive power.
\textbf{\color{red}Another issue is about handling distributed queries, where either selection criteria or selection values could come from distributed sources -- the biggest problem could be in the APIs they [DO NOT] provide}.

%TODO: Further it has to comply with dataset constraints

\subsection{Ease of use: Query Language over the mediated schema}
	Even a very simple structured query language that also contain entity names over the mediated schema may seem {\color{red}hard} to learn, especially in the beginning.
   On the other hand, at CERN the names in  the mediated schema are referring to real-world entities that {\color{red}are fairly consistently named} (even though there may exist slight differences in their naming on different data-sources). 
   So some sort of guiding helping user to build the query shall be useful.

   Supporting non-structured keyword queries is also worth investigation as many users reporting they are missing this Google-like search experience. {\color{red}Because DAS uses Boolean IR we could only rank specific structured queries.}

	Further a minimum set of search predicates is imposed by APIs (mainly because of performance reasons) and user has to be at least informed what is he expected to provide.

Also to consider: input ambiguity and typos, some queries are more common than others.

\begin{itemize}
	     					\item  Query prioritization: {\color{red}maybe we could have prioritization in general, the heavier your query is the more you have to wait in favour of  the light queries}: 
	     					(we could have some sort of query cost evaluation based on history or manually predefined scores per API)

                		\item Explore more intelligent caching

@@

					\item Could displaying Partial Results improve performance?
					\begin{itemize}
					     \item given current APIs to do aggregation over data sources, DAS has to fetch ALL records matching the query instead of only the first page.
%                        \item ??? do we have cases then multiple APIs of the same service could be called for the same data, can we choose just one???
                        \item {\color{red}e.g. intermediary aggregation results while still calculating}

@@

	                	 \item scale testing - if we are storing lots of historical info. MongoDB is not so performant if DB cant fit in memory. One of well known solutions is installing SSD.                        
\end{itemize}

\subsection{Handling distributed search efficiently}

\pagebreak
\include{proposed_solutions}

\pagebreak
\include{literature_review}

\section{Work status}
TODO

\begin{verbatim}
- obtained DB copy of biggest data provider DBS (currently 80 GB + 200 GB indexes)
- preliminary literature review (to be continued deeper + waiting for book arrival: Principles of Data Integration)

--- some analysis of DAS logs

Upcomming Work items:
- couple of fairly simple prototypes of UI/access patters for simpler DAS querying
- check on performance improvements after creation of materialized view(s)
- interview (more) DAS users
\end{verbatim}

@@

%\nocite{*}
\thispagestyle{empty}
\bibliography{refs}

\pagebreak
\appendix
\section{DAS Query logs}

non parsible queries: 98923

\subsection*{Most common query patterns (Feb 2012-June 2012)}
\begin{verbatim}
total valid queries: 569408

50.16% (285605) :	dataset dataset.name=?
13.54% (77071)  :	site dataset.name=?
8.96%  (51035) 	:	file dataset.name=?
5.88%  (33504) 	:	run dataset.name=?
2.59%  (14739) 	:	release dataset.name=?
2.11%  (12036) 	:	config dataset.name=?
1.65%  (9420) 	:	dataset run.run_number=?
1.34%  (7642) 	:	block dataset.name=?
1.24%  (7084) 	:	dataset site.name=?
1.15%  (6562) 	:	dataset dataset.name=? release.name=?
1.10%  (6287) 	:	parent dataset.name=?
0.96%  (5488) 	:	file file.name=?
0.92%  (5245) 	:	dataset dataset.name=? status.name=?
0.77%  (4363) 	:	dataset release.name=?
0.75%  (4257) 	:	file dataset.name=? run.run_number=?
0.68%  (3874) 	:	run run.run_number=?
0.53%  (2994) 	:	site site.name=?
0.45%  (2576) 	:	site file.name=?
0.45%  (2556) 	:	dataset file.name=?
0.43%  (2438) 	:	lumi file.name=?
0.40%  (2282) 	:	dataset dataset.name=? site.name=?
0.35%  (1999) 	:	file block.name=?
0.35%  (1970) 	:	dataset dataset.name=? run.run_number=?
0.29%  (1640) 	:	group dataset.name=?
0.29%  (1631) 	:	lumi run.run_number=?
0.25%  (1402) 	:	child dataset.name=?
0.22%  (1268) 	:	run file.name=?
0.21%  (1204) 	:	block block.name=?
0.19%  (1088) 	:	release release.name=?
0.16%  (936) 	:	site block.name=?
0.12%  (703) 	:	file dataset.name=? lumi.number=? run.run_number=?
0.10%  (594) 	:	parent file.name=?
0.08%  (455) 	:	dataset block.name=?
0.06%  (365) 	:	dataset dataset.name=? datatype.name=? release.name=?
0.06%  (360) 	:	file dataset.name=? site.name=?

Interesting non structured queries:
T!_CERN
*herwig*/AODSIM

\end{verbatim}
% py dataset=/DYJetsToLL_M-50_TuneZ2Star_8TeV-madgraph-tarball/Summer12_DR53X-PU_S10_START53_V7A-v1/AODSIM

>>>>

Using a highly structured language like SQL is problematic because users need to know not only the language but also where to find the information and also lots of technical details like schema. A data integration system based on simple structured queries is already in place. Various improvements including support for less restricted keyword queries, improvements to system's usability and performance still have to be researched.

@@

\section{Case study: the CMS Experiment at CERN}
Users' information need may vary greatly depending on their role, however most of the time they are interested in locating a \textit{full set} of entities matching some selection criteria, e.g.:
%\footnote{in contrary to exploratory approach where  a subset of best matches is enough}
   \begin{itemize}
  		\item find all \textit{files} from \textit{dataset(s)} matching wild-card query each containing some of the 'interesting' \textit{runs} from a list provided (Release validation teams)
         \item find (all) \textit{datasets} related some specific physics phenomena\footnote{In case of dataset this data is present in filename or run} together with conditions describing how this data was recorded by detector or simulated which are present in separate autonomous system than the datasets (Physicists)
         % TODO:is this a good example? conditions is a separate query now...
         \item find all \textit{datasets} matching some pattern stored at a given \textit{site} 
         	  \begin{small}(filtering attributes from separate services)\end{small}
   \end{itemize}                		
% In such dynamic environment it is preferred to choose local-as-view approach for information integration instead of creating a global schemata\cite{Koch00CERN}.

@@

The Data Aggregation System (DAS)\cite{Kuznetsov2010, Kuznetsov2011} was created which allows integrated access to a number of proprietary data-sources by processing user's queries on demand - it determines  data-sources are able to answer\footnote{This is done by a mapping between flat mediated schema ('DAS keys') into web-service methods and their arguments. Then system queries all services that could provide a result of expected type with given parameters}, queries them, merges the results and caches them for subsequent use. DAS uses \textit{Boolean retrieval model} as users are often interested in retrieving ALL the items matching their query.

Currently the queries specify what entity the user is interested in (dataset, file, etc) and provide selection criteria (attribute=value, name BETWEEN [v1, v2]) operators. The combined query results could be later 'piped' for further filtering, sorting or aggregation (min, max, avg, sum, count, median), e.g.:

{\small 
\begin{verbatim}
dataset=*RelVal* | grep dataset.nevents >1000 | avg(dataset.size), median(dataset.size)
\end{verbatim}
}

@@

\section{Problem statement}
%TODO: Further it has to comply with dataset constraints

\subsection{Ease of use}

For an IR system with wide variety of users, it important to provide an easy to use interface with fairly flat learning curve, while at the same time not loosing support of fairly complex queries.
%
	Even a simple structured query language containing entity names over the mediated schema may seem hard to learn%
	\footnote{On the other hand, at CERN the names in  the mediated schema are referring to real-world entities that {\color{red}are fairly consistently named} (even though there may exist slight differences in their naming on different data-sources).}%
	, especially in the beginning. Therefore, it shall be useful to guide new users interactively through the process of building of their query.
%   
   Supporting non-structured keyword queries is also worth investigation as quite many users reported they are missing this Google-like search experience.

% TODO:	Further a minimum set of search predicates is imposed by APIs (mainly because of performance reasons) and user has to be informed what extra is he expected to provide.

% TODO:	Even in structured query mode users are prone to typos, use either autocompletion where possible or suggest alternative queries.

@@

Therefore, it is important to balance between returning locally cached query results quickly   and between issuing queries to slow data-services to get up to date results.

\begin{itemize}
						\item Low data providers performance (e.g. on DBS some queries requires joining many large tables)
	     				\item At the moment all queries are put into one pool and has to wait until some threads are available. If a couple of heavy queries were submitted before hand, even very light queries would have to wait long. Explore more advanced Query prioritization 
	     					(e.g. we could have some query cost model based on history or predefined scores per API)

                		\item Currently result items are cached for a fairly short period of time (5min-1h) and then completely discarded, however many entities are not changing that often - Explore more intelligent caching

@@

					\item Could displaying Partial Results improve performance?
					\begin{itemize}
					     \item given current APIs to do aggregation over data sources, DAS has to fetch ALL records matching the query instead of only the first page.
%                        \item ??? do we have cases then multiple APIs of the same service could be called for the same data, can we choose just one???
                        \item 
                        \begin{small}
                        {\color{red}e.g. displaying intermediary results of aggregation while still calculating with statistical bounds -- however statistical bounds are much meaningful only if items are well snuffled/randomized -- people could stop the query if they see a non-sense}
                        \end{small}

@@

	              %\item {\color{red}scale testing - if we are storing lots of historical info. MongoDB is not so performant if DB cant fit in memory. One of well known solutions is installing SSD.                        }
\end{itemize}

\subsection{Other}
\subsubsection*{Handling distributed search efficiently}
{\color{red} People may be filtering data items where selection criteria could be present on two separate autonomous sources, meaning with the current APIs available the results would have to be fetched separately from each source and ones not contained in both filtered out only afterwards that is suboptimal.}

\subsubsection*{Generic connector access relational databases}
{\color{red}Generic connector able to access relational databases with minimal manual integration could be useful for integrating proprietary systems that were not yet integrated and there's no resources available to build APIs. A possible use-case could be the \textit{Prep} database, but this still has to be discussed.

Depending on needs it could be either of exploratory or API/Query Forms based approach, where fake APIs are defined in terms of SQL queries (or even better in a simplified form that would generate SQL by taking schema into account)}

\include{proposed_solutions}

\include{literature_review}

\section{Work status}

\begin{verbatim}
- obtained DB copy of biggest data provider DBS (currently 80 GB + 200 GB indexes)
- preliminary literature review (to be continued deeper)
- initial analysis of DAS logs

Upcomming Work items:
- couple of fairly simple prototypes of UI/access patters for simpler DAS querying
- check performance improvements after creation of materialized view(s)
- interview more DAS users
\end{verbatim}

@@

%\nocite{*}
\thispagestyle{empty}
\begin{small}
\bibliography{refs}
\end{small}

\pagebreak
\include{das_logs}
